---
title: "Constrained GLLVM from Scratch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Constrained_GLLVM-from-Scratch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# IGNORE THIS: IT IS WORK IN PROGRESS

This is a toy problem, to show how the code works.

https://www.r-inla.org/faq#h.pb4g1pwbmtli

```{r setup}
library(LatentINLA)

```

First, create a small data set. Too small to really do anything with other than see how it gets used.

```{r CreateData}
  Y.mat <- matrix(1:50, ncol=5)
  Y <- rnorm(nrow(Y.mat))
  NLVs <- 1
```

So there are 5 species and 10 sites. We will assume the data are counts, and use 1 latent variable. 

First we format the data:

```{r FormatData}
dat <- FormatDataFrameForLV(Y.mat)
```



```{r CreateLVs, eval=FALSE}
# create LV vectors
LVs <- lapply(1:NLVs, function(lv, dat) CreateLVIndices(dat=dat, nLVs=lv), dat=Y.mat)
names(LVs) <- paste0("lv", 1:NLVs)
LVs[[2]]
```

Now we can write the formula:

```{r CreateFormula, eval=FALSE}
(Formula <- formula(paste0("Y ~ " , CreateFormulaRHS(LVs=LVs))))
```

And finally fit the model. The results are rubbish, of course. But at least we have them.

```{r FitModel, eval=FALSE}

Data <- as.list(as.data.frame(LVs))
Data$Y <- dat

# fit the model
toyLVmodel = inla(Formula, data=Data, family = rep("poisson", ncol(Data$Y)))
summary(toyLVmodel)
```
